---
title: "PS5 markdown"
author: "Kwanpoonisa Mahakeetea"
date: "2/13/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Gapminder data

We use gapminder dataset, downloaded from [**https://www.gapminder.org/data/**](https://www.gapminder.org/data/), however, the data structure there is quire complex, please use the dataset provided on canvas (in files/data).\
The variables are:\
**name** country name\
**iso3** 3-letter country code\
**iso2** 2-letter country code\
**region** broad geographic region\
**sub-region** more precise region\
**intermediate-region**\
**time** year\
**totalPopulation** total population\
**GDP_PC** GDP per capita (constant 2010 US\$)\
**accessElectricity** Access to electricity (% of population)\
**agriculturalLand** Agricultural land (sq. km)\
**agricultureTractors** Agricultural machinery, tractors (count)\
**cerealProduction** Cereal production (metric tons)\
**feritilizerHa** Fertilizer consumption (kilograms per hectare of arable land)\
**fertilityRate** total fertility rate (births per woman)\
**lifeExpectancy** Life expectancy at birth, total (years)\
**childMortality** Mortality rate, under-5 (per 1,000 live births)\
**youthFemaleLiteracy** Literacy rate, youth female (% of females ages 15-24)\
**youthMaleLiteracy** Literacy rate, youth male (% of males ages 15-24)\
**adultLiteracy** Literacy rate, adult total (% of people ages 15 and above)\
**co2** CO2 emissions (kt)\
**greenhouseGases** Total greenhouse gas emissions (kt of CO2 equivalent)\
**co2_PC** CO2 emissions (metric tons per capita)\
**pm2.5_35** PM2.5 pollution, population exposed to levels exceeding WHO Interim Target-1 value 36ug/m3\
**battleDeaths** Battle-related deaths (number of people)\

# 1 Load and check data (5pt)

```{r}
library(tidyverse)
```

You first task is to do a very simple data check:

\*\*1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating).\*\*

**2. (2pt) Load data. How many rows/columns do we have?**

```{r}
gm <- read_delim("gapminder.csv.bz2")
```

```{r}
nrow(gm)
ncol(gm)
```

There are 13055 rows and 25 columns.

**3. (2pt) Print a small sample of data. Does it look OK?**

```{r}
head(gm)
```

The small printed sample of data looks okay.

------------------------------------------------------------------------

# 2 Descriptive statistics (15pt)

**1. (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.**

```{r}
gm %>%
  select(iso2, iso3, name) %>%
  summarise(n1 = n_distinct(iso2), n2 = n_distinct(iso3), n3 = n_distinct(name))
```

There are 249 countries listed for 2-letter codes, 253 countries listed for 3-letter codes, and 250 countries listed for the actual name.

**2. If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?**

    **(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?**

```{r}
gm %>%
  group_by(iso2) %>%
  summarise(num = n_distinct(iso2)) %>%
  arrange(desc(num))
```

**There is one NA in the iso2 column. Since "NA" wasn't filtered out, it was counted towards the names for each iso-2 code.**

    **(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.**

```{r}
gm %>%
  group_by(name) %>%
  select(iso3, name) %>%
  summarise(num = n_distinct(iso3)) %>%
  filter(num > 1)
```

**3. (2pt) What is the minimum and maximum year in these data?**

```{r}
gm %>%
  filter(!is.na(time)) %>%
  summarise(min = min(time), max = max(time))
```

The minimum year is 1960 and the maximum year is 2019.

# 3 CO2 emissions (30pt)

Next, let's analyze CO2 emissions.

**1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data?**

```{r}
gm %>%
  filter(is.na(co2)) %>%
  group_by(time) %>%
  count()
```

```{r}
gm %>%
  filter(is.na(co2_PC)) %>%
  group_by(time) %>%
  count()
```

The year 1960 has the most missing data for both co2 and co2_PC.

**2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.**

library("ggplot2")

```{r}
co2_emissions_overtime <- gm %>%
  filter(iso3 == "USA"| iso3 == "CHN"| iso3 == "IND"| iso3 == "ARG"| iso3 == "BEL") %>%
ggplot() +
  geom_line(aes(x = time, y = co2, col = name)) + geom_point(aes(x = time, y = co2, col = name))

print(co2_emissions_overtime)
```

**3. (5pt) Now let's analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of the same countries. What does this figure suggest?**

```{r}
co2_PC_overtime <- gm %>%
  filter(iso3 == "USA"| iso3 == "CHN"| iso3 == "IND"| iso3 == "ARG"| iso3 == "BEL") %>%
ggplot() +
  geom_line(aes(x = time, y = co2_PC, col = name)) + geom_point(aes(x = time, y = co2_PC, col = name))

print(co2_PC_overtime)
```

**4. (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see. Note: just compute averages over countries and ignore the fact that countries are of different size. Hint: Americas 2016 should be 4.80.**

```{r}
gm %>%
  filter(time %in% c(1960, 2016), !is.na(region), !is.na(co2_PC)) %>%
  group_by(time, region) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  print()
```

America had the highest average CO2 emissions per capita in 1960 and Europe had the highest average CO2 emssions per capita in 2016.

**5. (7pt) Make a barplot where you show the previous results--average CO2 emissions per capita across continents in 1960 and 2016. Hint: it should look something along these lines(**Please refer to the spec for the graph**)**

```{r}
gm %>%
  filter(time %in% c(1960, 2016), !is.na(region), !is.na(co2_PC)) %>%
  group_by(time, region) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  ggplot(aes(x = region, y = avg_co2_PC, fill = as.factor(time))) + geom_col(position = "dodge") + 
  labs(title = "Average CO2 Emissions Per Capita Across Continents",
       x = "Continent",
       y = "Average CO2 Emissions Per Capita") +
  scale_fill_discrete(name = "Year")
```

**6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent).**

# Largest CO2 emission countries:

## Americas:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Americas") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  head(3) %>%
  pull(name)
  
```

## Asia:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Asia") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  head(3) %>%
  pull(name)
```

## Africa:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Africa") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  head(3) %>%
  pull(name)
```

## Europe:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Europe") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  head(3) %>%
  pull(name)
```

## Oceania:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Oceania") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  head(3) %>%
  pull(name)
```

# Smallest CO2 emission countries:

## Americas:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Americas") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  tail(3) %>%
  pull(name)
  
```

## Asia:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Asia") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  tail(3) %>%
  pull(name)
```

## Africa:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Africa") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  tail(3) %>%
  pull(name)
```

## Europe:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Europe") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  tail(3) %>%
  pull(name)
```

## Oceania:

```{r}
gm %>%
  filter(!is.na(co2_PC)) %>%
  filter(region == "Oceania") %>%
  filter(time == "2016") %>%
  group_by(name) %>%
  summarise(avg_co2_PC = mean(co2_PC)) %>%
  arrange(avg_co2_PC) %>%
  tail(3) %>%
  pull(name)
```

------------------------------------------------------------------------

# 4 GDP per capita (50pt)

Let's look at GDP per capita (*GDP_PC*)

**1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better. Comment what do you see there.**

```{r}
gm %>%
  filter(time == "1960", !is.na(name), !is.na(GDP_PC), !is.na(lifeExpectancy)) %>%
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) + geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(1, 10)) +
    labs(title = "GDP Per Capita Versus Life Expectancy by Country in 1960",  
          x = "GDP per capita", 
          y = "life expectancy (year)")
```

**2. (4pt) Make a similar plot, but this time use 2019 data only.**

```{r}
gm %>%
  filter(time == "2019", !is.na(name), !is.na(GDP_PC), !is.na(lifeExpectancy)) %>%
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) + geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(1, 10)) +
    labs(title = "GDP Per Capita Versus Life Expectancy by Country in 2019",  
          x = "GDP per capita", 
          y = "life expectancy (year)")
```

**3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?**

From 1960 and 2019, Europe's life expectancy and GDP per capita have been higher than other regions in the world. Other regions have also progressively increased their GDP per capita and life expectancy. However, it can be seen that Africa's life expectancy and GDP per capita have remained around the same the past 60 years.

**4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures? Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.**

```{r}
gm %>%
  filter(time == "1960" | time == "2019") %>%
  filter(!is.na(region), !is.na(time), !is.na(lifeExpectancy)) %>%
  group_by(region, time) %>%
  summarise(avg_le = mean(lifeExpectancy)) %>%
  arrange(-avg_le)
```

Yes, it fits with the results from the previous questions.

\*\*5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see. Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\*\*

```{r}
gm %>%
  filter(time == "1960" | time == "2019") %>%
  filter(!is.na(region), !is.na(time), !is.na(lifeExpectancy)) %>%
  group_by(region, time) %>%
  summarise(avg_le = mean(lifeExpectancy)) %>%
  mutate(prev = lag(avg_le, default = 0), growth = avg_le - prev) 
```

Asia has the highest growth for 2019. However, I was unable to find prev and growth for 1960 because there's no data before that year.

**6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!**

```{r}
gm %>%
  filter(time == "1960"| time == "2019") %>%
  filter(!is.na(GDP_PC)) %>%
  group_by(time) %>%
  ggplot(aes(GDP_PC, fill = factor(time))) + geom_histogram(position = "dodge") + 
  labs(title = "GDP Per Capita for 1960 and 2019",
       x = "GDP per capita",
       y = "Count") +
  scale_fill_discrete(name = "time")

```

**7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.) Hint: check out the function rank()! Hint2: 17 for 1960.**

```{r}
gm %>%
  select(name, time, lifeExpectancy) %>% 
  filter(name != "", !is.na(time), !is.na(lifeExpectancy)) %>%
  filter(time == "1960") %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America")
```

```{r}
gm %>%
  select(name, time, lifeExpectancy) %>% 
  filter(name != "", !is.na(time), !is.na(lifeExpectancy)) %>%
  filter(time == "2019") %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America")
```

In 1960, the US's life expectancy was 17 and in 2019 it dropped down to 46.

**8. If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019--what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? Hint: 0.0904 for 1960.**

```{r}
gm %>%
  filter(time == "1960" | time == "2019") %>%
  filter(!is.na(region), !is.na(lifeExpectancy)) %>%
  mutate(le_rank = rank(desc(lifeExpectancy)), nc = n(), re_rank = le_rank/nc) %>% 
  select(name, le_rank, time, nc, re_rank) %>%
  filter(name == "United States of America")
```

------------------------------------------------------------------------

**Finally tell us how many hours did you spend on this PS.** I spent 12 hours.
